---
title: "Data Gateways: Accessing Public Hi-C Data Portals"
# output: github_document
editor_options: 
  chunk_output_type: inline
---

### Data Gateways: Accessing Public Hi-C Data Portals
The objective of this notebook is to develop working knowledge of accessing Hi-C datasets from:
1) 4DN Consortium
2) DNA Zoo Project

# Load packages and objects
```{r}
library(ggplot2)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(fourDNData)
library(DNAZooData)
library(rtracklayer)
```
## 4DN Data Portal: 4D Nucleome Data Coordination and Integration Center (DCIC)

Use fourDNData() function to access gateway to 4DN-hosted Hi-C files.
```{r}
library(fourDNData)
head(fourDNData())
```

# Query individual files with specific file parameters.
```{r}
cf <- fourDNData(experimentSetAccession = '4DNESJNPEKZD', type = 'mcool')
```

This will download and cache any queried files locally.
```{r}
cf
availableChromosomes(cf)
availableResolutions(cf)
import(cf, focus = "chr4:10000001-20000000", resolution = 5000)
```

To fetch a specific Hi-C file type, such as 'pairs', use the 'type' argument.
```{r}
pairs_f <- fourDNData(experimentSetAccession = '4DNESJNPEKZD', type = 'pairs') 
print(pairs_f)
import(pairs_f)
```

For type = 'insulation', this fetches a .bigwig track file. This track type corresponds to a genome-wide insulation score. Once fetched from the 4DN portal, local files can be imported in R.
```{r}
library(rtracklayer)
fourDNData(experimentSetAccession = '4DNES25ABNZ1', type = 'insulation') |> 
    import(as = 'Rle')
```

For type = 'boundaries', it fetches a .bed file which lists annotated borders between topological domains. This file type can also be imported in R and will generate a GRanges object.
```{r}
fourDNData(experimentSetAccession = '4DNES25ABNZ1', type = 'boundaries') |> 
    import()
```

For type = 'compartments', this will fetch a .bigwig track file. This track corresponds to a selected genome-wide eigenvector computed by cooltools, which represents A/B compartments.
```{r}
fourDNData(experimentSetAccession = '4DNES25ABNZ1', type = 'compartments') |> 
    import()
```

# Query complete experiment datasets
Instead of importing multiple files individually for each single experimentSet accession ID, we can import all available files simultaneously using fourDNHiCEperiment().
```{r}
hic <- fourDNHiCExperiment('4DNESJNPEKZD')
```
This aggregates all the components into a single HiCExperiment object with topologicalFeatures and metadata slots.
```{r}
metadata(hic)
```
## DNA Zoo
The DNA Zoo Consortium is a data portal maintained by a group who aims to correct and refine genome assemblies using Hi-C approaches. The DNAZooData() function from the similarly titled package provides a gateway to their Hi-C files.
```{r}
library(DNAZooData)
head(DNAZooData())
```
Fetching a Hi-C dataset generated from a tardigrade sample.
```{r}
hicfile <- DNAZooData(species = 'Hypsibius_dujardini')
hicfile
```
We can check the metadata in the HicFile parsed from the DNA Zoo portal.
```{r}
metadata(hicfile)$organism
```



