---
title: "Manipulating Hi-C data in R"
author: Vinson Zeng
date: March 29, 2024
output: github_document
---

The objective of this notebook is to develop working knowledge of:<br>
1. Modifying information associated with an existing HiCExperiment object<br>
2. Subsetting a HiCExperiment object<br>
3. Coercing a HiCExperiment object in a base data structure


#### Load packages and objects
```{r}
setwd("~/repos/Hi-C_workshop")
.libPaths("~/rlibrary")

library(ggplot2)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(HiContactsData)
```

#### Generate an example `hic` object 
The HiCExperiment object will be created from an example `.cool` file in the `HiContactsData` package.
```{r}
library(HiCExperiment)
library(HiContactsData)

#=== download example `.mcool` file and cache it locally
coolf <- HiContactsData('yeast_wt', 'mcool')

#=== create connection to disk-stored `.mcool` file
cf <- CoolFile(coolf)
cf

#=== import contacts from the long arm of chromosome `II`, at resolution `2000`
hic <- import(cf, focus = 'II:300001-813184', resolution = 2000)
hic
```

#### Subsetting a contact matrix

The two approaches to subset a Hi-C contact matrix include: <br>
1. Subsetting before importing - leverages random access disk-stored contact matrix to only import interactions overlapping with a genomic locus of interest.
2. Subsetting after importing - parse entire contact matrix in memory, subset interactions overlapping with a genomic locus of interest.

#### Subset before import with `focus`:
```{r}
import(cf, focus = 'II:300001-800000', resolution = 2000)
```

Subset to off-diagonal genomic location using pairs of coordinates query:
```{r}
import(cf, focus = 'II:300001-400000|II:600001-700000', resolution = 2000)
```

Subset interactions constrained within single chromosome:
```{r}
import(cf, focus = 'II', resolution = 2000)
```

Subset interactions to retain those between two chromosomes:
```{r}
import(cf, focus = 'II|III', resolution = 2000)
```

Subset interactions to retain those between parts of two chromosomes:
```{r}
import(cf, focus = 'II:300001-800000|V:1-500000', resolution = 2000)
```

#### Subsetting after import with `subsetByOverlaps` or `[`
```{r}
telomere <- GRanges("II:700001-813184")
subsetByOverlaps(hic, telomere) |> interactions()
```

Subsetting with `[`:
```{r}
hic["II:800001-813184"]
```

Subset off-diagonal genomic location using pairs of coordinates query:
```{r}
hic["II:300001-320000|II:800001-813184"]
```

Subsetting for vectors of several chromosomes cannot be performed with `focus` from disk-stored data. This scenario requires `[`-based in-memory subsetting of pre-imported data.
```{r}
hic[c('II', 'III', 'IV')]
```

#### Zooming in a `HiCExperiment`

The term "zooming" refers to dynamically changing the resolution of a HiCExperiment. The `zoom` function only works for multi-resolution contact matrices (e.g. `.mcool` or `.hic`) and it does not change the `focus`. It only affects the `resolution` and `interactions`.
```{r}
hic
zoom(hic, 4000)
```

#### Updating a `HiCExperiment` object

Immutable `HiCExperiment` slots:<br>
<ul>
  <li> `fileName(hic)` </li>
  <li> `focus(hic)` </li>
  <li> `resolutions(hic)` </li>
  <li> `interactions(hic)` </li>
</ul>

Mutable `HiCExperiment` slots:<br>
<ul>
  <li> `scores(hic)` </li>
  <li> `topologicalFeatures(hic)` </li>
  <li> `pairsFile(hic)` </li>
  <li> `metadata(hic)` </li>
</ul>

```{r}
# example of mutable slots for a HiCExperiment object

#=== create additional topologicalFeatures or modify existing ones with topologicalFeatures()<-
topologicalFeatures(hic, 'CTCF') <- GRanges(c(
    "II:340-352", 
    "II:3520-3532", 
    "II:7980-7992", 
    "II:9240-9252" 
))
topologicalFeatures(hic, 'CTCF')

topologicalFeatures(hic, 'loops') <- GInteractions(
    topologicalFeatures(hic, 'CTCF')[rep(1:3, each = 3)],
    topologicalFeatures(hic, 'CTCF')[rep(1:3, 3)]
)
topologicalFeatures(hic, 'loops')

#=== adding pairsFile after importing ContactFile into a HiCExperiment object
pairsf <- HiContactsData('yeast_wt', 'pairs.gz')
pairsFile(hic) <- pairsf

#=== update metadata for a HiCExperiment
metadata(hic) <- list(
    info = "HiCExperiment created from an example .mcool file from `HiContactsData`", 
    date = date()
)
metadata(hic)
```

#### Coercing `HiCExperiment` objects

Coercing functions allow for transformation of data stored as a `HiCExperiment` into another class. These include:.<br>
<ul>
  <li>`as.matrix()`</li>
  <li>`as.data.frame()`</li>
</ul>

```{r}
#=== `as.matrix` coerces into dense matrix by default
as.matrix(hic) |> class()

as.matrix(hic) |> dim()

#=== specify scores to be used when coercing into a matrix
as.matrix(hic, use.scores = "balanced")[1:5, 1:5]

as.matrix(hic, use.scores = "count")[1:5, 1:5]

#=== for a spare matrix, include sparse = TRUE
as.matrix(hic, use.scores = "count", sparse = TRUE)[1:5, 1:5]
```

Coercing `interactions` into a data frame (rectangular).
```{r}
as.data.frame(hic) |> head()

```

