---
title: "Hi-C Data Visualization"
author: Vinson Zeng
date: March 29, 2024
output: github_document
---

The objective of this notebook is to review working knowledge the visualization tools in R offered by:<br>
1) `HiContacts`<br>
2) `HiCExperiment`

#### Load packages and objects
```{r}
library(ggplot2)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(HiContactsData)
library(HiContacts)
library(rtracklayer)
```

#### Single map

Disk-stored Hi-C contact matrices can be visualized by importing the interactions of interest over a genomic location into a `HiCExperiment` object, followed by using `plotMatrix` from `HiContacts`.
```{r}
#=== generate example hic object
# create HiCExperiment object from example .cool file in HiContactsData package

# download `.mcool` file and cache locally
coolf <- HiContactsData('yeast_wt', 'mcool')

# create connection to disk-stored `.mcool` file
cf <- CoolFile(coolf)
cf

# import contacts from chromosome V at a resolution of 2000
hic <- import(cf, focus = 'V', resolution = 2000)

# plot the matrix
plotMatrix(hic)
```

#### Horizontal map

